**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-08-01 submission/manuscript.Rmd
fix a couple of sentences
before Pat's edits
update

Advances in sequencing technology and decreases in the costs of generating 16S rRNA gene sequences have allowed rapid exploration of the human microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and there is an increasing demand for methods that identify associations between microbiome members and human health. However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. They learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. The high-stake task of predicting the correct diagnosis with high confidence relies on a ML model that is built with rigorous methods. However, the field's use of ML lacks clarity and consistency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019]. More notably, we commonly see flawed ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate predictive performance or reporting few or only the best outcomes of cross-validation. Even when there are seperate testing sets to evaluate predictive performance, there are large differences between cross-validation and testing performances and large confidence intervals in testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. These practices prevent the development of generalizable models, where the model makes accurate predictions with newly acquired data as well as it does with the training data.

Moreover, there is a lack of discussion on why a particular ML model is utilized and a lack of emphasis on the strengths and weaknesses of it. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretability is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. These models are also called black box ML models because they are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction. These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018]. The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].

The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibility. We need to strive toward better machine learning practices by (1) implementing rigorous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of predictive performance, complexity, interpretability and computational efficiency. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of CRC patients. The human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN). We established modeling pipelines for three linear models with different forms of regularization; L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear kernel. We also developed four non-linear models; SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared the predictive performance, interpretability and computational efficiency of the seven models to highlight the importance of model selection. We established standards for ML pipeline construction, predictive performance evaluation and model interpretation for microbiome-associated ML models that have a binary prediction task.

@@

__Model selection and pipeline construction__

We first determined the dataset and the ML models to use in our study. We used a previously published study on a CRC cohort of 490 patients with 261 cases of SRN. For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (normal or SRN colorectal lesions as defined by colonoscopies). We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost).

We used ML models with different classification algorithms and regularization parameters. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero). Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGBoost, as well as an SVM with non-linear kernel.

We established a ML pipeline where we train and validate each of the seven models [Figure 1]. We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets. Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.

Hyperparameters are the rules that are learned from the training data in a classification algorithm. When they are tuned over a full grid search and selected by validation, they make the ML model better are predicting. We selected hyperparameter settings using repeated five-fold cross-validation on the training/validation set [Figure 1]. We chose the best hyperparameter setting for each model based on its predictive performance on the validation set using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance. Similar to the initial data-split, five-fold cross-validation for hyperparameter selection was also stratified to maintain the overall label distribution on the training and validation sets. The cross-validation of each hyperparameter setting was repeated over 100 randomizations to get a robust reading of predictive performance.

We then trained the full training/validation dataset with the selected hyperparameters. We used the held-out test set to evaluate the predictive performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].

@@

In this study, we also established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. First of all, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data. In all the models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Furthermore, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model. We showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits. The randomization and resampling of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics. We also performed a full grid search for hyperparameter settings when training our ML models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default. And finally, we used the AUROC metric in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.

@@

  The colorectal health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas. The study had 261 normal and 229 SRN samples. The bacterial abundances are the features used to predict colonic health of the patients. Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size of our smallest sample and were subsampled at the same distances. They were then transformed by scaling to a [0-1] range.

>>>>

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease. However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods.  However,  ML methodology in the microbiome field has flaws that have not been addressed previously. There is a lack of transparency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019]; models are being developed and evaluated without a seperate held-out test data [@flemer_oral_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @montassier_pretreatment_2016; @ai_systematic_2017]; there is large variation between the cross-validation outcomes [@dai_multi-cohort_2018; @ai_systematic_2017] and between cross-validation and testing performances [@mossotto_classification_2017] . These practices limit reproducibility, cause overoptimism for model performance and prevent the development of generalizable models, where the model makes accurate predictions with newly acquired data as well as it does with the training data. Nevertheless, the microbiome field is making progress to avoid potential pitfalls of ML. More and more studies are now validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and they are introducing frameworks to accurately use ML tools [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].

Among microbiome researchers, there has been a trend towards using more complex ML models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @@wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. There is an implicit assumption that the more complex models are better because they are more complex. Although these models may be better at incorporating non-linear relationships or yield better predictions, they are also called black box ML models because they are not inherently interpretable. These models require post hoc explanations to quantify the importance of each feature in making a prediction. Depending on the application of the model researchers may chose to use different modeling approaches. For example, a researcher trying to identify the populations causing a disease would likely want a more interpretable model whereas a clinician may emphasize performance. Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018].

The lack of transparency in model selection and using flawed modelling and interpretation methods negatively impacts model validity. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches using the same dataset and  pipeline. We established modeling pipelines for three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also developed four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared the predictive performance, interpretability and computational efficiency of the seven models to highlight the importance of model selection. To demonstrate the performance of these modelling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of samples (N=490) connected to clinically significant disease where there is ample evidence that it is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we established standards for ML pipeline construction, evaluated predictive performance, and demonstrated model interpretation for models. This framework can be easily applied to other host-associated and environmental microbiome problems.

@@

__Model selection and pipeline construction__ We established a ML pipeline where we train and validate each of the seven models [Figure 1]. We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the data-split was stratified to maintain the overall label distribution in both the training and test sets. For example, our example dataset represented data from 490 individuals, 261 had normal colons and 229 had a screen relevant neoplasia (SRN). After the data-split, the training set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training data was used for hyperparameter selection and the test set was used for evaluating predictive performance.

We used ML models with different classification algorithms and regularization methods. Regularization is a technique to discourage overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero). Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGBoost, as well as an SVM with non-linear kernel.

Hyperparameters are the rules that are learned from the training data in a classification algorithm. For example regularization term (C) that decides how big the penalty for overfitting will be, is an hyperparameter. The model becomes more generalizable when we tune for the optimal C value. Similar to regularization term C, other hyperparameters can be tuned over a full grid search and selected by validation to build better models. We selected hyperparameter settings by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. Similar to the initial data-split, five-fold CV was also stratified to maintain the overall label distribution.  We chose the best hyperparameter setting for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance. The cross-validation of each hyperparameter setting was repeated over 100 randomizations to get a robust reading of predictive performance.

We then trained the full training dataset with the selected hyperparameters. We used the held-out test set to evaluate the testing predictive performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].

@@

In this study, we also established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. First of all, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data. In all the models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Furthermore, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model. We showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits. The randomization and resampling of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics. Additionally, we performed a full grid search for hyperparameter settings when training our ML models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default. And finally, we used the AUROC metric in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.

@@

  The colorectal health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas. The study had 261 normal and 229 SRN samples. The bacterial abundances are the features used to predict colorectal health of the patients. For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (normal or SRN colorectal lesions as defined by colonoscopies). We established modeling pipelines for a binary prediction task Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size of our smallest sample and were subsampled at the same distances. They were then transformed by scaling to a [0-1] range.
